Location and Scale-Invariant Power Transformations for Transforming Data to

### 2025-03-21 manuscript_rmarkdown/manuscript.Rmd
Update p-values for empirical tests.

---
title: "Location and Scale-Invariant Power Transformations for Transforming Data to
  Normality"
author: "Alex Zwanenburg, Steffen Löck"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: lualatex
    keep_tex: true
papersize: a4
bibliography: refs.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{booktabs}
- \DeclareMathOperator*{\argmax}{argmax}
- \DeclareMathOperator*{\argmin}{argmin}
- \DeclareMathOperator{\sgn}{sgn}
---

@@

```{=latex}
\begin{equation}
f_{\alpha}(x; \mu, \sigma, \beta) = \frac{2 \alpha \left(1 - \alpha\right)}{\sigma}
\begin{cases}
f_{\beta}\left( \left(1 - \alpha \right) \frac{\left| x - \mu \right|}{\sigma} \right) & \text{, } x \leq \mu \\
f_{\beta}\left( \alpha \frac{\left| x - \mu \right|}{\sigma} \right) & \text{, } x > \mu
\end{cases}
\end{equation}
```
Here, $\alpha \in (0,1)$ is a skewness parameter. $\alpha > 0.5$ creates a distribution with a negative skew, i.e. a left-skewed distribution. A right-skewed distribution is created for $\alpha < 0.5$. $\mu \in \mathcal{R}$ and $\sigma \in (0, \infty)$ are location and scale parameters, respectively. $f_{\alpha}$ thus describes the probability density function of an asymmetric generalised normal distribution, which we will refer to here and parametrise as $\mathcal{AGN}\left(\mu, \sigma, \alpha, \beta \right)$.

@@

```{=latex}
\begin{equation}
F_{\alpha}^{-1}(p; \mu, \sigma, \beta) =
\begin{cases}
\mu + \frac{\sigma}{1 - \alpha} F_{\beta}^{-1} \left( \frac{p}{2 \alpha}\right) & \text{, } p \leq \alpha \\
\mu + \frac{\sigma}{\alpha} F_{\beta}^{-1} \left( \frac{1 + p - 2 \alpha}{2 \left(1 - \alpha \right)} \right) & \text{, } p > \alpha
\end{cases}
\end{equation}
```
The quantile function for the asymmetric generalised normal distribution $F_{\alpha}^{-1}$ thus incorporates the quantile function $F_{\beta}^{-1}$ of the symmetric generalised normal distribution. $F_{\beta}^{-1}$ was derived by Griffin to be [@Griffin2018-bf]:

```{=latex}
\begin{equation}
F_{\beta}^{-1}(p) = \sgn\left(p - 0.5 \right) F_{\Gamma}^{-1}\left(2 \left|p - 0.5 \right|; 1 / \beta \right)
\end{equation}
```
Here, $F_{\Gamma}^{-1}$ is the quantile function of the gamma distribution with shape $1 / \beta$, which can be numerically approximated.

## Empirical central normality test

Power transformations aim to transform features to a normal distribution.
However, this may not always be successful or possible.
Deviations from normality can be detected by normality tests, such as the Shapiro-Wilk test [@Shapiro1965-zd].
In practice, normality tests may be too stringent with large sample sizes, outliers, or both.
Here we develop an empirical test for central normality.
The null hypothesis $H_0$ is that the distribution is centrally normal.
The alternative hypothesis $H_1$ is that the distribution is not centrally normal.

Let central normality be defined as the normality of central portion $\kappa$ of the data, i.e. $\mathbf{X}_{\text{central}} = \left\{x_i \in \mathbf{X} \, | \,  \frac{1-\kappa}{2} \leq  p_i \leq \frac{1 + \kappa}{2}\right\}$, with $p_i$ probabilities of the empirical distribution, as previously.
We then compute the residual errors between the z-scores of the transformed feature $\phi^{\lambda, x_0, s} (\mathbf{X})$ and the expected z-scores from a standard normal distribution: $r_i =\left| \left( \phi^{\lambda, x_0, s}(x_i) - \mu_M)\right) / \sigma_M - F^{-1}_{\mathcal{N}}(p_i) \right|$, with $\mu_M$ and $\sigma_M$ robust Huber M-estimates of location and scale of the transformed feature $\phi^{\lambda, x_0, s} (\mathbf{X})$ [@Huber1981-su].
The set of residual errors for the central portion of the data is then $\mathbf{R}_{\text{central}} = \left\{ r_i \in \left\{ r_1, r_2, \ldots, r_n\right\} \, | \,  \frac{1-\kappa}{2} \leq  p_i \leq \frac{1 + \kappa}{2}\right\}$.

The test statistic $\tau_{\text{ecn}}$ is then defined as:
```{=latex}
\begin{equation}
\tau_{\text{ecn}} = \frac{\sum_{i=1}^{N} r_i \left[r_i \in \mathbf{R}_{\text{central}}\right]}{\sum_{i=1}^N \left[r_i \in \mathbf{R}_{\text{central}}\right]} 
\end{equation}
```
Here $[\quad]$ denotes an Iverson bracket.
The test statistic is equal to the mean of the residual errors of the central portion of the data.

To use the test statistic, the central portion of the data needs to be defined and the Type 1 error rates determined.
We will do so in the [Simulation](#simulation) section.

@@

usethis::use_data(
  gof_lookup_table, 
  two_sided_function_parameters, 
  internal = TRUE, 
  overwrite = TRUE)
```

@@

## Empirical central normality test

To develop an empirical test for central normality we need to consider two parameters: the central portion $\kappa$ as a fixed parameter, and test statistic $\tau_{\text{ecn}}$. We will first define the central portion $\kappa$.

First we draw $m_d=10000$ random asymmetric generalised normal distributions. As before, each distribution is parametrised with a randomly chosen skewness parameter $\alpha \sim U\left(0.01, 0.99\right)$ and shape parameter $\beta \sim U\left(1.00, 5.00 \right)$. Location and scale parameters are set as $\mu = 0$ and $\sigma = 1$, respectively. $n = \lceil 10^\gamma \rceil$ values are then randomly drawn, with $\gamma \sim U\left(1.47, 3.00\right)$, which leads to between $30$ and $1000$ values being drawn to create $\mathbf{X}_i$. As before, for each distribution, up to $10 \%$ of instances are replaced by outlier values, and this is repeated $10$ times. We then compute residuals after optimising robust location- and scale-invariant transformations with the empirical tapered cosine weighting method.

Figure \ref{fig:empirical-central-normality-residual-error} shows the residual errors of the transformed distribution as function of the empirical probability.
As expected, the largest deviations from normality appear on the extremities of this range.
For Box-Cox transformations, the effect of outliers on the lower end of the distribution leads to noticeable asymmetry.
Between very low and very high empirical probabilities, residual errors are constrained and relatively flat.
This is the candidate range for the central portion of the data.

```{r empirical-central-normality-residual-error, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=3.0}
cap <- paste0(
  "Residual error as a function of empirical probability. ",
  "Percentiles of the error are shown for robust, location- and scale-invariant transformations of 10000 randomly drawn asymmetric ",
  "generalised normal distributions, for each of which outliers were randomly drawn 10 times ",
  "(up to 10% of samples). Larger errors occur at the edges of each distribution."
)

.plot_residuals(
  plot_theme = plot_theme,
  manuscript_dir = manuscript_dir
)

```
We then consider the empirical probability of type I errors: the probability of incorrectly classifying a centrally normal distribution as being non-normal based on the test statistic $\tau_{\text{ecn}}$.
Under the assumption that the asymmetric generalised normal distributions are centrally normal after robust transformation, this produces the relationship shown in Figure \ref{fig:empirical-central-normality-type-1-error-rate}.
Since for $\kappa \leq 0.80$ error curves for both transformation methods are similar, we fix the value of the central portion $\kappa$ to 0.80.
In the remaining we will use the empirical central normality test statistic values defined using robust location- and scale-invariant Yeo-Johnson transformations, as it is more conservative (see Appendix E).
This leads to test statistic values listed in Table \ref{tab:empirical-central-normality}.

```{r empirical-central-normality-type-1-error-rate, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=3.0}
cap <- paste0(
  "Type 1 error rate of transformed asymmetric generalised normal distributions as function of ",
  "the test statistic $\\tau_{\\text{ecn}}$ for the central portion $\\kappa$ of the distribution.")

.plot_type_1_error_rate(
  plot_theme = plot_theme,
  manuscript_dir = manuscript_dir)

@@

```{r include = FALSE, eval = FALSE}
# This writes the lookup table for the empirical central normality test to the
# power.transform package.
load(file = "./R/sysdata.rda")

gof_lookup_table <- .get_test_statistics_data(
  manuscript_dir = manuscript_dir,
  reduce = TRUE,
  as_table = FALSE)
gof_lookup_table[, "method" := NULL]

usethis::use_data(
  gof_lookup_table, 
  two_sided_function_parameters, 
  internal = TRUE, 
  overwrite = TRUE)
```

```{r include = FALSE, eval = FALSE}
# This generates the data for the table below.
table_data <- .get_test_statistics_data(manuscript_dir = manuscript_dir)
```

```{=latex}
\begin{table}
\begin{center}
\caption{Test statistic $\tau_{\text{ecn}}$ for empirical central normality at $\kappa = 0.80$ as a function of Type I error rate.}
\label{tab:empirical-central-normality}
\begin{tabular}{l | c c c c c c c}

\toprule
type I error rate & 0.50 & 0.20 & 0.10 & 0.05 & 0.02 & 0.01 & 0.001 \\

\midrule
$\tau_{\text{ecn}}$ & 0.041 & 0.062 & 0.075 & 0.088 & 0.103 & 0.115 & 0.154 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}
```

In Figure \ref{fig:empirical-central-normality-examples} we apply the empirical central normality test to assess central normality of features that are composed of a mixture of samples drawn from two normal distributions. With increased separation of the underlying normal distributions, the probability of the feature being centrally normal decreases, as expected.

```{r empirical-central-normality-examples, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=2.5}
cap <- paste0(
  "Bi-modal distributions and empirical central normality test results. ",
  "The feature (black) is a mixture of two identical sample sets (blue and orange) ",
  "drawn from normal distributions that are offset by a distance $d$. ",
  "We use the empirical centrally normality test to compute the probability for the hypothesis that ",
  "the distribution is centrally normal. As may be observed, with increasing offset $d$ ",
  "the probability that the feature is centrally normal decreases. ",
  "Quantile-quantile plots are drawn below each distribution."
)

.plot_bimodal_distribution_test(
  plot_theme = plot_theme,
  manuscript_dir = manuscript_dir)

@@

data <- .plot_experimental_invariance(plot_theme = plot_theme, method = "yeo_johnson")
data$p
```

@@

A common feature in health-related datasets is age.
Here we use data on 228 patients with lung cancer that was collected and published by Loprinzi et al. [@Loprinzi1994-cd].
The age in the cohort was $62.4 \pm 9.1$ (mean ± standard deviation) years.
Applying conventional and invariant Yeo-Johnson transformations to patient age yielded the following results, see Figure \ref{fig:experimental-results-invariance}:
no transformation (sum of residuals with normal distribution $\sum r_i = 16.5$);
conventional transformation ($\lambda = 2.0$, $\sum r_i = 11.5$, $\mu_{YJ} = 1.8 \cdot 10^3$, $\sigma_{YJ} = 0.5 \cdot 10^3$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 2.0$, $\sum r_i = 11.5$, $\mu_{YJ} = 1.8 \cdot 10^3$, $\sigma_{YJ} = 0.5 \cdot 10^3$);
location- and scale-invariant transformation ($\lambda = 0.9$, $\sum r_i = 8.8$, $\mu_{YJ} = 1.2$, $\sigma_{YJ} = 1.1$);
and robust location- and scale-invariant transformation ($\lambda = 0.8$, $\sum r_i = 10.6$, $\mu_{YJ} = 1.2$, $\sigma_{YJ} = 1.0$).

@@

Gorman, Williams and Fraser recorded body mass (in grams) of 342 penguins of three different species [@Gorman2014-eo].
The body mass was $(4.2 \pm 0.8) \cdot 10^3$ (mean ± standard deviation) grams, and not centrally normal ($p = 0.03$).
Applying conventional and invariant Yeo-Johnson transformations to body mass yielded the following results, see Figure \ref{fig:experimental-results-invariance}:
no transformation (residual sum $\sum r_i = 48.0$);
conventional transformation ($\lambda = -0.5$, $\sum r_i = 32.2$, $\mu_{YJ} = 2.1$, $\sigma_{YJ} = 4 \cdot 10^{-3}$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = -0.5$, $\sum r_i = 32.2$, $\mu_{YJ} = 2.1$, $\sigma_{YJ} = 4 \cdot 10^{-3}$);
location- and scale-invariant transformation ($\lambda = 0.5$, $\sum r_i = 26.8$, $\mu_{YJ} = 0.9$, $\sigma_{YJ} = 0.9$);
and robust location- and scale-invariant transformation ($\lambda = 0.4$, $\sum r_i = 23.1$, $\mu_{YJ} = 0.8$, $\sigma_{YJ} = 0.8$).

@@

Geospatial datasets usually contain coordinates.
The Ames housing dataset contains data on 2930 properties that were sold between 2006 and 2010 [@De-Cock2011-jf], including their geospatial coordinates.
The latitude was $42.03 \pm 0.02)$ (mean ± standard deviation).
Applying conventional and invariant Yeo-Johnson transformations to latitude yielded the following results, see Figure \ref{fig:experimental-results-invariance}:
no transformation (residual sum $\sum r_i = 328$);
conventional transformation ($\lambda = 62.1$, $\sum r_i = 319$, $\mu_{YJ} = 4.8 \cdot 10^{99}$, $\sigma_{YJ} = 0.1 \cdot 10^{99}$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 95.4$, $\sum r_i = 319$, $\mu_{YJ} = 6.4 \cdot 10^{153}$, $\sigma_{YJ} = 0.3 \cdot 10^{153}$);
location- and scale-invariant transformation ($\lambda = 1.5$, $\sum r_i = 326$, $\mu_{YJ} = -1.2$, $\sigma_{YJ} = 0.8$);
and robust location- and scale-invariant transformation ($\lambda = 1.4$, $\sum r_i = 311$, $\mu_{YJ} = -1.3$, $\sigma_{YJ} = 0.9$).

@@

data <- .plot_experimental_outlier_robustness(plot_theme = plot_theme, method = "yeo_johnson")
data$p
```

@@

The Top Gear dataset contains data on 297 vehicles that appeared on the BBC television show *Top Gear* [@Alfons2021-kc].
Within this dataset, the fuel consumption feature contains outliers due to highly fuel-efficient vehicles.
Applying conventional and invariant Yeo-Johnson transformations to the fuel consumption feature yielded the following results, see Figure \ref{fig:experimental-results-outlier-robustness}:
no transformation (residual sum $\sum r_i = 54$, $p=0.76$);
conventional transformation ($\lambda = -0.1$, $\sum r_i = 55$, $\mu_{YJ} = 3.0$, $\sigma_{YJ} = 0.3$, $p=0.01$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 0.8$, $\sum r_i = 48$, $\mu_{YJ} = 29$, $\sigma_{YJ} = 15$, $p=0.55$);
location- and scale-invariant transformation ($\lambda = -1.3$, $\sum r_i = 44$, $\mu_{YJ} = 0.5$, $\sigma_{YJ} = 0.1$, $p=0.03$);
and robust location- and scale-invariant transformation ($\lambda = -0.9$, $\sum r_i = 50$, $\mu_{YJ} = 2.0$, $\sigma_{YJ} = 2.3$, $p=0.58$).

Outliers cause non-robust transformations to fail to transform the data to a centrally normal distribution (empirical central normality test $p = 0.01$ and $p=0.03$ for conventional and invariant transformations, respectively).
Robust transformations produce distributions that are centrally normal (empirical central normality test $p > 0.05$).

@@

The ischemic stroke dataset contains historic data from 126 patients with risk at ischemic stroke [@Kuhn2019-kt].
These patients underwent Computed Tomography Angiography to characterize the carotid artery blockages.
Angiography imaging was then assessed, and various characteristics related to the blood vessels and the disease are measured.
The maximum arterial wall thickness feature contains several instances with outlier values.
Applying conventional and invariant Yeo-Johnson transformations to this feature yielded the following results, see Figure \ref{fig:experimental-results-outlier-robustness}:
no transformation (residual sum $\sum r_i = 110$, $p=0.56$);
conventional transformation ($\lambda = -0.7$, $\sum r_i = 30$, $\mu_{YJ} = 1.0$, $\sigma_{YJ} = 0.1$, $p=0.01$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 1.1$, $\sum r_i = 136$, $\mu_{YJ} = 7.2$, $\sigma_{YJ} = 14$, $p=0.61$);
location- and scale-invariant transformation ($\lambda = 0.2$, $\sum r_i = 12$, $\mu_{YJ} = -11.8$, $\sigma_{YJ} = 6.9$, $p=0.13$);
and robust location- and scale-invariant transformation ($\lambda = -0.6$, $\sum r_i = 27$, $\mu_{YJ} = 0.7$, $\sigma_{YJ} = 0.1$, $p=0.10$).

Non-robust transformations failed to produce a centrally normal distribution (empirical central normality test $p=0.01$ and $p=0.02$ for conventional and invariant transformations, respectively).
Robust transformations produce distributions that are centrally normal (empirical central normality test $p > 0.05$).

>>>>

---
title: "Location and Scale-Invariant Power Transformations for Transforming Data to
  Normality"
author: "Alex Zwanenburg, Steffen Löck"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: lualatex
    keep_tex: true
papersize: a4
bibliography: refs.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{amsthm}
- \usepackage{booktabs}
- \DeclareMathOperator*{\argmax}{argmax}
- \DeclareMathOperator*{\argmin}{argmin}
- \DeclareMathOperator{\sgn}{sgn}
---

@@

\begin{equation}
f_{\alpha}(x; \mu, \sigma, \beta) = \frac{2 \alpha \left(1 - \alpha\right)}{\sigma}
\begin{cases}
f_{\beta}\left( \left(1 - \alpha \right) \frac{\left| x - \mu \right|}{\sigma} \right) & \text{, } x \leq \mu \\
f_{\beta}\left( \alpha \frac{\left| x - \mu \right|}{\sigma} \right) & \text{, } x > \mu
\end{cases}
\end{equation}

Here, $\alpha \in (0,1)$ is a skewness parameter. $\alpha > 0.5$ creates a distribution with a negative skew, i.e. a left-skewed distribution. A right-skewed distribution is created for $\alpha < 0.5$. $\mu \in \mathcal{R}$ and $\sigma \in (0, \infty)$ are location and scale parameters, respectively. $f_{\alpha}$ thus describes the probability density function of an asymmetric generalised normal distribution, which we will refer to here and parametrise as $\mathcal{AGN}\left(\mu, \sigma, \alpha, \beta \right)$.

@@

\begin{equation}
F_{\alpha}^{-1}(p; \mu, \sigma, \beta) =
\begin{cases}
\mu + \frac{\sigma}{1 - \alpha} F_{\beta}^{-1} \left( \frac{p}{2 \alpha}\right) & \text{, } p \leq \alpha \\
\mu + \frac{\sigma}{\alpha} F_{\beta}^{-1} \left( \frac{1 + p - 2 \alpha}{2 \left(1 - \alpha \right)} \right) & \text{, } p > \alpha
\end{cases}
\end{equation}

The quantile function for the asymmetric generalised normal distribution $F_{\alpha}^{-1}$ thus incorporates the quantile function $F_{\beta}^{-1}$ of the symmetric generalised normal distribution. $F_{\beta}^{-1}$ was derived by Griffin to be [@Griffin2018-bf]:

\begin{equation}
F_{\beta}^{-1}(p) = \sgn\left(p - 0.5 \right) F_{\Gamma}^{-1}\left(2 \left|p - 0.5 \right|; 1 / \beta \right)
\end{equation}

Here, $F_{\Gamma}^{-1}$ is the quantile function of the gamma distribution with shape $1 / \beta$, which can be numerically approximated.

## Central normality test

Power transformations aim to transform features to a normal distribution.
However, this may not always be successful or possible.
Deviations from normality can be detected by normality tests, such as the Shapiro-Wilk test [@Shapiro1965-zd].
In practice, normality tests may be too stringent with large sample sizes, outliers, or both.
Here we develop a test for central normality.

\begin{definition}[central portion of a sequence]
Let $\mathbf{X} = \left\{x_1, x_2, \ldots, x_n | x_i \in \mathbb{R}\right\}$ be a finite sequence of length $n > 0$,
ordered so that $x_1 \leq x_2 \leq \ldots \leq x_n$.
Let $p_i = \frac{i - 1/3}{n + 1/3}$, with $i = 1, 2, \ldots n$ be the percentile value
corresponding to each element in $\mathbf{X}$. For elements with tied values in $\mathbf{X}$,
percentile values are replaced by the average in their group, i.e.
if $x_j = x_{j+1} = \ldots = x_{j+m}$, then 
$p_j' = p_{j+1}' = \ldots = p_{j+m}' = 1/(m+1)\sum_j^{j+m}p_j$.
Furthermore, let central portion $\kappa \in (0,1)$.

Then the central portion of $\mathbf{X}$ is 
$\mathbf{X}_{\text{c}} = \left\{x_i \in \mathbf{X} \, | \,  \frac{1-\kappa}{2} \leq  p_i \leq \frac{1 + \kappa}{2}\right\}$.

\end{definition}

\begin{definition}[Residual errors of the central portion of a sequence]

Let the residual error for each element of $\mathbf{X}$ be $r_i =\left| \frac{x_i - \mu_M}{\sigma_M} - F^{-1}_{\mathcal{N}}(p_i) \right|$, 
with $\mu_M$ and $\sigma_M$ robust Huber M-estimates of location and scale of $\mathbf{X}$ [@Huber1981-su], 
and $F^{-1}_{\mathcal{N}}$ the quantile function of the normal distribution $\mathcal{N}(0, 1)$.

Then, the set of residual errors of the central portion of $\mathbf{X}$ (i.e. $\mathbf{X}_{\text{c}}$) is
$\mathbf{R}_{\text{c}} = \left\{ r_i \in \left\{ r_1, r_2, \ldots, r_n\right\} \, | \,  \frac{1-\kappa}{2} \leq  p_i \leq \frac{1 + \kappa}{2}\right\}$.

\end{definition}

\begin{definition}[Central normality of a sequence]

The central portion of a sequence is normally distributed if the sum of residual
errors is equal to zero:
$\sum_{r_j \in \mathbf{R}_{\text{c}}} r_j = 0$

\end{definition}

In practice, a finite sequence sampled from a normal distribution $\mathcal{N}(\mu, \sigma)$ will have a non-zero sum of residual errors.

\begin{definition}[Central normality test]
The null-hypothesis is defined as: $\mathcal{H}_0: \sum_{r_j \in \mathbf{R}_{\text{c}}}r_j = 0$

THe alternative hypothesis is defined as: $\mathcal{H}_1: \sum_{r_j \in \mathbf{R}_{\text{c}}}r_j > 0$
\end{definition}

The null-hypothesis $\mathcal{H}_0$ is that the central portion of sequence $\mathbf{X}$
is normally distributed, with the alternative hypothesis $\mathcal{H}_1$ that it
is not normally distributed. To test this hypothesis, a test statistic is computed
and compared against a critical value.

\begin{definition}[Central normality test statistic]
Let $\alpha$ be the significance level for the central normality test.
Let $m$ be the number of elements of $\mathbf{R}_{\text{c}}$

The test statistic is then $\tau_{n, \kappa} = \frac{1}{m}\sum_{r_j \in \mathbf{R}_{\text{c}}}r_j$.

\end{definition}

The null-hypothesis should be rejected at significance level $\alpha$ if 
$\tau_{n, \kappa} \geq \tau_{\alpha, n, \kappa, \text{critical}}$.

The central portion of the data needs to be defined and the Type 1 error rates determined to provide critical test statistics.
We will do so in the [Simulation](#simulation) section.

@@

usethis::use_data(
  central_normality_lookup_table,
  empirical_central_normality_lookup_table,
  two_sided_function_parameters, 
  internal = TRUE, 
  overwrite = TRUE
)
```

@@

## Central normality and empirical central normality test

To develop a test for central normality we need to consider two
parameters: the central portion $\kappa$ as a fixed parameter, and test
statistic $\tau_{\text{ecn}}$. We will first define the central portion
$\kappa$.

For each number of samples $n \in \left\{\lfloor 10^\nu \rfloor | \nu \in \left\{0.7500, 0.8125, \ldots, 4.0000 \right \} \right\}$
we randomly draw $m_d = 30000$ ordered sequences $\mathbf{X}$ from $\mathcal{N}(0,1)$.
This dataset is used to compute the critical test statistic values for the central normality test.
Additionally, we add $10 \%$ outliers by randomly replacing elements of each ordered sequence.
The dataset with outliers is used to compute the critical test statistic values for the empirical variant of the central normality test.
For each sequence we then compute $\tau_{n, \kappa}$ for $\kappa \in \left\{0.60, 0.70, 0.80, 0.90, 0.95, 1.00\right\}$.

The dataset with outliers is used to select a suitable value for $\kappa$.
Figure \ref{empirical-central-normality-kappa} shows $\tau_{\alpha = 0.95, n, \kappa}$
as a function of $n$ for different values of $\kappa$. With decreasing
$\kappa$, the test statistic curve decreases. The curves
for $\kappa = 0.60$ and $\kappa = 0.70$ are similar, whereas curves for 
$\kappa \geq 0.90$ are affected by outliers, as can be observed by comparing curves
of the central normality test with those of the empirical variant.

Since the (empirical) central normality tests assesses whether the central part of a sequence is normally distributed,
we use $\kappa = 0.70$. Lookup tables and curves for critical statistic values for central normality and empirical central
normality tests are shown in Appendix E.

```{r empirical-central-normality-kappa, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=2.5}
cap <- paste0(
  "Critical test statistic $\\tau_{\\alpha = 0.95, n, \\kappa}$ of the (empirical) ",
  "central normality test as function of $n$ for several values of central portion $\\kappa$. ",
  "The critical test statistics for central normality test are determined using fully normal data, ",
  "whereas the statistics for the empirical variant are determined using centrally normal data, ",
  "i.e. with fully normal data where 10% of elements are replaced by outliers. ",
  "emp: empirical"
)

.plot_test_statistic_centrality(
  plot_theme = plot_theme,
  manuscript_dir = manuscript_dir
)

@@

In Figure \ref{fig:empirical-central-normality-examples} we apply the empirical
central normality test to assess central normality of features that are composed
of a mixture of samples drawn from two normal distributions ($n = 100$ each).
With increased separation of the underlying normal distributions, the probability of the
feature being centrally normal decreases, as expected.

```{r empirical-central-normality-examples, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=2.5}
cap <- paste0(
  "Bi-modal distributions and empirical central normality test results. ",
  "The feature (black) is a mixture of two identical sample sets (blue and orange, $n = 100$) ",
  "drawn from normal distributions that are offset by a distance $d$. ",
  "We use the empirical centrally normality test to compute the probability for the hypothesis that ",
  "the distribution is centrally normal. As may be observed, with increasing offset $d$ ",
  "the probability that the feature is centrally normal decreases. ",
  "Quantile-quantile plots are drawn below each distribution."
)

.plot_bimodal_distribution_test(
  plot_theme = plot_theme,
  manuscript_dir = manuscript_dir
)

@@

data <- .plot_experimental_invariance(plot_theme = plot_theme, method = "yeo_johnson")
data$plot
```

@@

A common feature in health-related datasets is age.
Here we use data on 228 patients with lung cancer that was collected and published by Loprinzi et al. [@Loprinzi1994-cd].
The age in the cohort was $62.4 \pm 9.1$ (mean ± standard deviation) years.
Applying conventional and invariant Yeo-Johnson transformations to patient age yielded the following results, see Figure \ref{fig:experimental-results-invariance}:
no transformation (sum of residuals with normal distribution $\sum r_i = 16.5$, $p = 0.71$);
conventional transformation ($\lambda = 2.0$, $\sum r_i = 11.5$, $\mu_{YJ} = 1.8 \cdot 10^3$, $\sigma_{YJ} = 0.5 \cdot 10^3$, $p = 0.94$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 2.0$, $\sum r_i = 11.5$, $\mu_{YJ} = 1.8 \cdot 10^3$, $\sigma_{YJ} = 0.5 \cdot 10^3$, $p = 0.94$);
location- and scale-invariant transformation ($\lambda = 0.9$, $\sum r_i = 8.8$, $\mu_{YJ} = 1.2$, $\sigma_{YJ} = 1.1$, $p = 0.95$);
and robust location- and scale-invariant transformation ($\lambda = 0.8$, $\sum r_i = 10.6$, $\mu_{YJ} = 1.2$, $\sigma_{YJ} = 1.0$, $p = 0.80$).

@@

Gorman, Williams and Fraser recorded body mass (in grams) of 342 penguins of three different species [@Gorman2014-eo].
The body mass was $(4.2 \pm 0.8) \cdot 10^3$ (mean ± standard deviation) grams, and not centrally normal ($p = 0.03$).
Applying conventional and invariant Yeo-Johnson transformations to body mass yielded the following results, see Figure \ref{fig:experimental-results-invariance}:
no transformation (residual sum $\sum r_i = 48.0$, $p < 0.001$);
conventional transformation ($\lambda = -0.5$, $\sum r_i = 32.2$, $\mu_{YJ} = 2.1$, $\sigma_{YJ} = 4 \cdot 10^{-3}$, $p = 0.06$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = -0.5$, $\sum r_i = 32.2$, $\mu_{YJ} = 2.1$, $\sigma_{YJ} = 4 \cdot 10^{-3}$, $p = 0.06$);
location- and scale-invariant transformation ($\lambda = 0.5$, $\sum r_i = 26.8$, $\mu_{YJ} = 0.9$, $\sigma_{YJ} = 0.9$, $p = 0.18$);
and robust location- and scale-invariant transformation ($\lambda = 0.4$, $\sum r_i = 23.1$, $\mu_{YJ} = 0.8$, $\sigma_{YJ} = 0.8$, $p = 0.40$).

@@

Geospatial datasets usually contain coordinates.
The Ames housing dataset contains data on 2930 properties that were sold between 2006 and 2010 [@De-Cock2011-jf], including their geospatial coordinates.
The latitude was $42.03 \pm 0.02)$ (mean ± standard deviation).
Applying conventional and invariant Yeo-Johnson transformations to latitude yielded the following results, see Figure \ref{fig:experimental-results-invariance}:
no transformation (residual sum $\sum r_i = 328$, $p < 0.001$);
conventional transformation ($\lambda = 62.1$, $\sum r_i = 319$, $\mu_{YJ} = 4.8 \cdot 10^{99}$, $\sigma_{YJ} = 0.1 \cdot 10^{99}$, $p < 0.001$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 95.4$, $\sum r_i = 319$, $\mu_{YJ} = 6.4 \cdot 10^{153}$, $\sigma_{YJ} = 0.3 \cdot 10^{153}$, $p < 0.001$);
location- and scale-invariant transformation ($\lambda = 1.5$, $\sum r_i = 326$, $\mu_{YJ} = -1.2$, $\sigma_{YJ} = 0.8$, $p < 0.001$);
and robust location- and scale-invariant transformation ($\lambda = 1.4$, $\sum r_i = 311$, $\mu_{YJ} = -1.3$, $\sigma_{YJ} = 0.9$, $p < 0.001$).

@@

data <- .plot_experimental_outlier_robustness(plot_theme = plot_theme, method = "yeo_johnson")
data$plot
```

@@

The Top Gear dataset contains data on 297 vehicles that appeared on the BBC television show *Top Gear* [@Alfons2021-kc].
Within this dataset, the fuel consumption feature contains outliers due to highly fuel-efficient vehicles.
Applying conventional and invariant Yeo-Johnson transformations to the fuel consumption feature yielded the following results, see Figure \ref{fig:experimental-results-outlier-robustness}:
no transformation (residual sum $\sum r_i = 54$, $p=0.58$);
conventional transformation ($\lambda = -0.1$, $\sum r_i = 55$, $\mu_{YJ} = 3.0$, $\sigma_{YJ} = 0.3$, $p < 0.001$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 0.8$, $\sum r_i = 48$, $\mu_{YJ} = 29$, $\sigma_{YJ} = 15$, $p=0.32$);
location- and scale-invariant transformation ($\lambda = -1.3$, $\sum r_i = 44$, $\mu_{YJ} = 0.5$, $\sigma_{YJ} = 0.1$, $p < 0.001$);
and robust location- and scale-invariant transformation ($\lambda = -0.9$, $\sum r_i = 50$, $\mu_{YJ} = 2.0$, $\sigma_{YJ} = 2.3$, $p=0.35$).

Outliers cause non-robust transformations to fail to transform the data to a centrally normal distribution (empirical central normality test $p < 0.001$ for conventional and invariant transformations).
Robust transformations produce distributions that are centrally normal (empirical central normality test $p > 0.05$).

@@

The ischemic stroke dataset contains historic data from 126 patients with risk at ischemic stroke [@Kuhn2019-kt].
These patients underwent Computed Tomography Angiography to characterize the carotid artery blockages.
Angiography imaging was then assessed, and various characteristics related to the blood vessels and the disease are measured.
The maximum arterial wall thickness feature contains several instances with outlier values.
Applying conventional and invariant Yeo-Johnson transformations to this feature yielded the following results, see Figure \ref{fig:experimental-results-outlier-robustness}:
no transformation (residual sum $\sum r_i = 110$, $p=0.86$);
conventional transformation ($\lambda = -0.7$, $\sum r_i = 30$, $\mu_{YJ} = 1.0$, $\sigma_{YJ} = 0.1$, $p=0.02$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 1.1$, $\sum r_i = 136$, $\mu_{YJ} = 7.2$, $\sigma_{YJ} = 14$, $p=0.91$);
location- and scale-invariant transformation ($\lambda = 0.2$, $\sum r_i = 12$, $\mu_{YJ} = -11.8$, $\sigma_{YJ} = 6.9$, $p=0.23$);
and robust location- and scale-invariant transformation ($\lambda = -0.6$, $\sum r_i = 27$, $\mu_{YJ} = 0.7$, $\sigma_{YJ} = 0.1$, $p=0.18$).

The conventional non-robust transformation failed to produce a centrally normal distribution (empirical central normality test $p=0.02$).
Robust transformations produce distributions that are centrally normal (empirical central normality test $p > 0.05$).
